---
title: "White Flight and School Racial Segregation"
shorttitle: "School Segregation"
keywords: [segregation, white flight]
thanks: Thanks to everyone for checking this out.
reference-section-title: References
bibliography: ../bibliography/project.bib
csl: ../bibliography/chicago-parenthetical.csl
format:
  aog-article-pdf:
    keep-tex: false
    include-in-header: 
      text: |
        \usepackage{dcolumn}
#  submittable-pdf:
#     keep-tex: false
#     fig-pos: "!t"
#     include-in-header: 
#       text: |
#         \usepackage{dcolumn}
  submittable-docx: default
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
load(here("data", "data_constructed", "analytical_data.RData"))

```

# Introduction and Background

Despite racially segregated schools being considered unconstitutional since the Supreme Court decision of Brown v. Board 1954, the education system in the U.S.is still heavily racially segregated. White flight is a phenomena that describes the pattern of a large amount of white families leaving urban areas into suburban areas. Typically, white families have the socioeconomic resources to be able to move middle-class or upper middle-class suburban areas. White families also have negative expectations of Black and Hispanic schools, and decide not to send their children there. In other words, White families prefer to send their children to white schools, to avoid the negative stereotypes that they
believe exist in Black and Hispanic schools. With the existence of private schools and the socioeconomic resources that white families have, they are able to enroll their children in private schools, without necessarily needing to move to a different city or neighborhood. 

This project looks at the relationship between the racial 
segregation of private schools and public schools. The research question 
is does the (n) amount of private schools correlate to the increase 
in racially segregated schools by school district in the U.S.? I'm interested in looking at if there are less private schools for white families to access, will there be less racially segregated public schools?

My unit of analysis would be counties because private schools are not a part of school districts. The outcome or dependent variable would be levels of segregation (if a population is majority-non-white or majority-white) and the independent variable would be the number of private schools in a county. 



# Data and Methods

There were two data sets that I used: National Center for Education Statistics 
(2019-2020) and American Community Surveys 5-year Estimates (ACS 2017-2021). 
The NCES data includes information on private and public schools from the years 
2019 to 2020 and the ACS data includes control variables on the county level 
from the years 2017 to 2021. The variables used from the private school and 
public schools are the county FIPS codes, the number of private schools, the 
racial demographics of each public school, and the number of public schools. 
The control variables used from the ACS data set were the median household income,
the highest level of educational attainment, and the population density on the 
county level. The independent variable is the number of private schools and the 
dependent variable is the dissimilarity index of white students and non-white 
students in public schools. My hypothesis was the more private schools there are 
in a county, the higher the segregation or dissimilarity between white and 
non-white students in public schools. 


# Results

To look at the relationship between the number of private schools and public
school dissimilarity, I created a scatter plot which is shown in @fig-1. This 
figure shows that this relationship is not linear and is heavily right-skewed. 
Just looking at the first figure, there seems to be a light positive relationship
between the number of private schools and the racial dissimilarity of public
schools, but it's a small amount and this graph is not the best representation
of this relationship. The second graph I created in @fig-2 shows cone-shaped 
data with fitted-values and residuals. This means that there is an IID violation 
and problems of heteroskedasticity. Again, there is not a strong linear 
relationship. 

To attempt to fix the IID violation I used log transformations on the 
independent variable in @fig-3. However, using the number of private schools
as the independent variable is not the best variable to use to capture a 
relationship with dissimilarity. There seems to be a slightly positive 
association, but it is still cone-shaped (heteroskedasticity) and not linear.
Having the percentage of all schools that are private or the percentage of 
all students who go to private schools would be better independent variables. 

For the models in @tbl-reg, in model 1 there is a positive relationship
with the n_private coefficient, which means the more private schools in a 
county, the higher the public school dissimilarity. However, The only 
statistically significant outcome is in model 1 and not in models 2 or 3. For 
the control variables the relationships are positive, but very small and not
statistically significant. As stated before, calculating a different 
independent variable that will still capture the relationship between
private schools and public school segregation would provide better results. 



# Conclusions

This project looked at racial school segregation and private schools at the 
county level in the U.S. Survey data from NCES and ACS were used to calculate
the relationship between the number of private schools in a county and the
racial dissimilarity of white and non-white students in public schools. The data
showed that there was a positive relationship between the two variables, meaning
the more private schools in a county the higher the dissimilarity index of
public schools. However, the independent variable (number of private schools)
was not the best way to represent this relationship, leading to non-linear
IID violation of heteroskedasticity. The data was heavily skewed to the right, 
not displaying the relationship as best as it could with a different independent
variable. Future work on this project would calculate the percentage of 
all schools that are private and the percentage of all students who go to 
private schools. It will also have better models with different control variables. 
```{r}
#| label: fig-1
#| fig-cap: Relationship Between the number of Private Schools and Public School Dissimilarity 

ggplot(school_dissim, aes(x=n_private, y=dissim))+
  geom_point()+
  geom_smooth(se = FALSE)+
  labs(x = "number of private schools", y = "public school dissimilarity", subtitle = "priavte schools and racial dissimilarity")+
  theme_bw()
```


```{r}
#| label: fig-2
#| fig-cap: Private schools and Racial Dissimilarity - residual vs. fitted values

model <- lm(dissim~n_private, data = school_dissim)
augment(model)

ggplot(augment(model), aes(x = .fitted, y = .resid))+
  geom_point()+
  geom_hline(yintercept = 0, linetype = 2)+
  geom_smooth(se = FALSE)+
  labs(x = "dissimilarity fitted values", y = "model residuals")+
  theme_bw()
```



```{r}
#| label: fig-3
#| fig-cap: log transformation

ggplot(school_dissim, aes(x = n_private, y = dissim))+
  geom_jitter(alpha = 0.5)+
  geom_smooth(method = "lm", color = "red", se = FALSE)+
  scale_x_log10()+
  labs(x = "number of private schools", y = "public school dissimilarity")+
  theme_bw()
```


```{r}
#| label: tbl-reg
#| tbl-cap: Relationship between number of private schools and public school racial dissimilarity by County
#| results: asis

model1 <- lm(dissim~n_private+median_household_income, data=school_dissim)
 
model2 <- lm(dissim~n_private+median_household_income+total_pop, data=school_dissim)
 
model3 <- lm(dissim~n_private+median_household_income+total_pop+pop_density, data=school_dissim)


texreg(list(model1, model2, model3),
       digits = 3,
       caption.above=TRUE, 
       include.rsquared=TRUE,
       include.adjrs=FALSE,
       include.nobs=TRUE,
       include.rmse=FALSE)

```
